Robust Co-occurrence Quantification for Lexical Distributional Semantics

### 2016-06-07 figures/interaction-cds.tex
Matt
Mentoring comments

\begin{figure*}
  \centering
    \centerline{\includegraphics[width=1.11\textwidth]{supplement/figures/SimLex999-interaction-cds}}
    \caption{\textbf{The effect of context probability estimation and smoothing on SimLex-999.}}
    \label{fig:interaction-cds}
\end{figure*}

>>>>

\begin{figure*}
  \centering
    \centerline{\includegraphics[width=1.11\textwidth]{supplement/figures/SimLex999-interaction-cds}}
    \caption{{\bf Effect of PMI variant (\texttt{discr}), smoothing (\texttt{cds}) and frequency weighting (\texttt{freq}) on SimLex-999.} Error bars correspond to a 95\% confidence interval as the value is estimated by aggregating over omitted parameters: \texttt{neg} and similarity.}
    \label{fig:interaction-cds}
\end{figure*}

### 2016-06-07 figures/interaction-neg.tex
Lexical tasks
Figures
Update
Pre-submission.
Mentoring comments

\begin{figure*}
  \centering
    \includegraphics[width=\textwidth]{supplement/figures/SimLex999-interaction-cds}
    \caption{\textbf{Context distribution smoothing.}}
    \label{fig:interaction-cds}
\end{figure*}

>>>>

\begin{figure*}
  \centering
    \centerline{\includegraphics[width=1.11\textwidth]{supplement/figures/SimLex999-interaction-neg}}
    \caption{\textbf{The behaviour of shifted PMI (SPMI) on SimLex-999.} \texttt{discr=spmi}, \texttt{freq=1} and \texttt{neg=1} corresponds to positive PMI.}
    \label{fig:interaction-neg}
\end{figure*}

### 2016-06-07 paper.tex
Mentoring comments

\aclfinalcopy % Uncomment this line for the final submission
\def\aclpaperid{30} %  Enter the acl Paper ID here

@@

\maketitle
\begin{abstract}
Previous optimisations of parameters affecting the word-context association measure used in distributional vector space models have focused either on high-dimensional vectors with hundreds of thousands of dimensions, or dense vectors with dimensionality of few hundreds; but dimensionality of few thousands is often applied in compositional tasks as it is still computationally feasible and does not require the dimensionality reduction step. We present a systematic study of the interaction of the parameters of the association measure and vector dimensionality, and derive parameter selection heuristics that achieve performance across word similarity and relevance datasets competitive with the results previously reported in the literature achieved by highly dimensional or dense models.
\end{abstract}

@@

An alternative way of obtaining a vector space with few dimensions, usually with just 100--500, is the use of SVD as a part of Latent Semantic Analysis \cite{ARIS:ARIS1440380105} or another models such as SGNS \cite{mikolov2013efficient} and GloVe \cite{pennington2014glove}. However, these models take more time to instantiate in comparison to weighting of a co-occurrence matrix and bring more parameters to explore. In this work we focus on vector spaces that directly weight a co-occurrence matrix and report results for SVD, GloVe and SGNS from the study of \newcite{TACL570} for comparison.

The mismatch of experiments with non-dense models in vector dimensionality between lexical and compositional tasks gives rise to a number of questions:
\begin{compactitem}
\item To what extent does model performance depend on vector dimensionality?
\item Do parameters influence 200K and 1K dimensional models similarly? Can the findings of \newcite{TACL570} be directly applied to models with few thousand dimensions?
\item If not, can we derive suitable parameter selection heuristics which take account of dimensionality?
\end{compactitem}

To answer these questions, we perform a systematic study of distributional models with a rich set of parameters on SimLex-999  \cite{hill2014simlex}, a lexical \emph{similairty} dataset, and test selected models on MEN \cite{Bruni:2014:MDS:2655713.2655714}, a lexical \emph{relatedness} dataset.

@@

Most co-occurrence weighting schemes in distributional semantics are based on \emph{point-wise mutual information} (PMI, Equation~\ref{eq:pmi}) (see e.g.~\newcite{J90-1003}, \newcite{Turney:2010:FMV:1861751.1861756}, \newcite{NIPS2014_5477}).
%
\begin{equation}
  \label{eq:pmi}
  \operatorname{PMI}(x, y) = \log\frac{P(x,y)}{P(x)P(y)}
\end{equation}
%
%PMI in its raw form is problematic: non-observed co-occurrences lead to infinite PMI values, making it impossible to compute similarity. A common ``fix'' to this problem is to replace all infinities with zeroes, and we use PMI hereafter to refer to a weighting with this fix.
%
As commonly done, we replace the infinite PMI values with zeroes and use PMI hereafter to refer to a weighting with this fix.
%
An alternative solution is to increment the probability ratio by 1; we refer to this as \textit{compressed PMI} (CPMI):
%
\begin{equation}
  \label{eq:cpmi}
  \operatorname{CPMI}(x, y) = \log\Big( 1 + \frac{P(x,y)}{P(x)P(y)} \Big)
\end{equation}

@@

Many approaches use only \emph{positive} PMI values, as  negative PMI values may not positively contribute to model performance \cite{Turney:2010:FMV:1861751.1861756}. This can be generalised to an additional cutoff parameter $k$ (\texttt{neg}) following \newcite{TACL570}, giving our third PMI variant (abbreviated as SPMI):
%
\begin{equation}
  \label{eq:ppmi}
  \operatorname{SPMI_k} = \max (0, \operatorname{PMI}(x, y) - \log k)
\end{equation}
%
We can apply the same idea to CPMI:
%
\begin{equation}
  \label{eq:pcpmi}
  \operatorname{SCPMI_k} = \max (0, \operatorname{CPMI}(x, y) - \log 2k)
\end{equation}

@@

As context words we select the 1K, 2K, 3K, 5K, 10K, 20K, 30K, 40K and 50K most frequent lemmatised nouns, verbs, adjectives and adverbs. All context words are part of speech tagged, but we do not distinguish between refined word types (e.g.~intransitive vs.~transitive versions of verbs).

@@

Overall, $\log n$ weighting should be used with PMI, CPMI and SCPMI. High-dimensional SPMI models show the same behaviour, but if $D<$ 10K, no weighting should be applied. SPMI and SCPMI should be preferred over CPMI and PMI. As Figure~\ref{fig:global-best-simlex} shows, our heuristics give performance close to the optimum for any dimensionality, with a large improvement over both an average parameter setting and the parameters suggested by \cite{TACL570} in a high-dimensional setting.\footnote{Our results using \cite{TACL570}'s parameters differ slightly from theirs due to different window sizes (5 vs 2).}

@@

In general, model performance depends on vector dimensionality (the best setup with 50K dimensions is better than the best setup with 1K dimensions by 0.03 on SimLex-999). Spaces with few thousand dimensions benefit from being dense and unsmoothed  ($k < 1$, global context probability); while high-dimensional spaces are better sparse and smooth  ($k > 1$, $\alpha = 0.75$). However, for unweighted and $n$-weighted models, these heuristics do not guarantee the best possible result because of the high variance of the corresponding scores. Based on this we suggest to use \logNSPMI/ or \logNSCPMI/ with dimensionality of at least 20K to ensure good performance on lexical tasks.

>>>>

% \aclfinalcopy % Uncomment this line for the final submission
\def\aclpaperid{30} %  Enter the acl Paper ID here

@@

\maketitle
\begin{abstract}
Previous optimisations of parameters affecting the word-context association measure used in distributional vector space models have focused either on high-dimensional vectors with hundreds of thousands of dimensions, or dense vectors with dimensionality of a few hundreds; but dimensionality of a few thousands is often applied in compositional tasks as it is still computationally feasible and does not require the dimensionality reduction step. We present a systematic study of the interaction of the parameters of the association measure and vector dimensionality, and derive parameter selection heuristics that achieve performance across word similarity and relevance datasets competitive with the results previously reported in the literature achieved by highly dimensional or dense models.
\end{abstract}

@@

An alternative way of obtaining a vector space with few dimensions, usually with just 100--500, is the use of SVD as a part of Latent Semantic Analysis \cite{ARIS:ARIS1440380105} or another models such as SGNS \cite{mikolov2013efficient} and GloVe \cite{pennington2014glove}. However, these models take more time to instantiate in comparison to weighting of a co-occurrence matrix, bring more parameters to explore and produce vector spaces with uninterpretable dimensions. In this work we focus on vector spaces that directly weight a co-occurrence matrix and report results for SVD, GloVe and SGNS from the study of \newcite{TACL570} for comparison.

The mismatch of experiments with non-dense models in vector dimensionality between lexical and compositional tasks gives rise to a number of questions:
\begin{compactitem}
\item To what extent does model performance depend on vector dimensionality?
\item Do parameters influence 200K and 1K dimensional models similarly? Can the findings of \newcite{TACL570} be directly applied to models with a few thousand dimensions?
\item If not, can we derive suitable parameter selection heuristics which take account of dimensionality?
\end{compactitem}

To answer these questions, we perform a systematic study of distributional models with a rich set of parameters on SimLex-999  \cite{hill2014simlex}, a lexical \emph{similairty} dataset, and test selected models on MEN \cite{Bruni:2014:MDS:2655713.2655714}, a lexical \emph{relatedness} dataset. Both datasets are currently widely used and surpass datasets stemming from information retrieval, WordSim-353 \cite{2002:PSC:503104.503110}, and computational linguistics, RG65 \cite{Rubenstein:1965:CCS:365628.365657}, in quantity by having more entries and quality by attention to evaluated relation.

@@

Most co-occurrence weighting schemes in distributional semantics are based on \emph{point-wise mutual information} (PMI, see e.g.~\newcite{J90-1003}, \newcite{Turney:2010:FMV:1861751.1861756}, \newcite{NIPS2014_5477}):
%
\begin{equation}
  \label{eq:pmi}
  \operatorname{PMI}(x, y) = \log\frac{P(x,y)}{P(x)P(y)}
\end{equation}
%
As commonly done, we replace the infinite PMI values,\footnotemark{} which arise when $P(x, y) = 0$, with zeroes and use PMI hereafter to refer to a weighting with this fix.

\footnotetext{We assume that the probability of a single token is always greater than zero as it appears in the corpus at least once.}

An alternative solution is to increment the probability ratio by 1; we refer to this as \textit{compressed PMI} (CPMI, see e.g.~\newcite{McGregorEtAl15JAGI}):
%
\begin{equation}
  \label{eq:cpmi}
  \operatorname{CPMI}(x, y) = \log\Big( 1 + \frac{P(x,y)}{P(x)P(y)} \Big)
\end{equation}

By incrementing the probability ratio by one, the PMI values from the segment of $(-\infty$; 0], when the joint probability $P(x,y)$ is less than the chance $P(x)P(y)$, are compressed into the segment of $(0; 1]$.
As the result, the space does not contain negative values, but has the same sparsity as the space with PMI values.
% In our experiments with spaces containing 56 thousand target words,

@@

Many approaches use only \emph{positive} PMI values, as  negative PMI values may not positively contribute to model performance and sparser matrices are more computationally tractable \cite{Turney:2010:FMV:1861751.1861756}. This can be generalised to an additional cutoff parameter $k$ (\texttt{neg}) following \newcite{TACL570}, giving our third PMI variant (abbreviated as SPMI):
%
\begin{equation}
  \label{eq:ppmi}
  \operatorname{SPMI_k} = \max (0, \operatorname{PMI}(x, y) - \log k)
\end{equation}
%
When $k = 1$ SPMI is equivalent to positive PMI. $k > 1$ increases the underlying matrix sparsity by keeping only highly associated co-occurrence pairs. $k < 1$ decreases the underlying matrix sparsity by including some unassociated co-occurrence pairs, which are usually excluded due to unreliability of probability estimation \cite{Dagan:1993:CWS:981574.981596}.

We can apply the same idea to CPMI:
%
\begin{equation}
  \label{eq:pcpmi}
  \operatorname{SCPMI_k} = \max (0, \operatorname{CPMI}(x, y) - \log 2k)
\end{equation}

@@

As context words we select the 1K, 2K, 3K, 5K, 10K, 20K, 30K, 40K and 50K most frequent lemmatised nouns, verbs, adjectives and adverbs. All context words are part of speech tagged, but we do not distinguish between refined word types (e.g.~intransitive vs.~transitive versions of verbs) and do not perform stop word filtering.

@@

Overall, $\log n$ weighting should be used with PMI, CPMI and SCPMI. High-dimensional SPMI models show the same behaviour, but if $D<$ 10K, no weighting should be applied. SPMI and SCPMI should be preferred over CPMI and PMI. As Figure~\ref{fig:global-best-simlex} shows, our heuristics give performance close to the optimum for any dimensionality, with a large improvement over both an average parameter setting and the parameters suggested by \cite{TACL570} in a high-dimensional setting.\footnote{Our results using \newcite{TACL570}'s parameters differ slightly from theirs due to different window sizes (5 vs 2).}

@@

In general, model performance depends on vector dimensionality (the best setup with 50K dimensions is better than the best setup with 1K dimensions by 0.03 on SimLex-999). Spaces with a few thousand dimensions benefit from being dense and unsmoothed  ($k < 1$, global context probability); while high-dimensional spaces are better sparse and smooth  ($k > 1$, $\alpha = 0.75$). However, for unweighted and $n$-weighted models, these heuristics do not guarantee the best possible result because of the high variance of the corresponding scores. Based on this we suggest to use \logNSPMI/ or \logNSCPMI/ with dimensionality of at least 20K to ensure good performance on lexical tasks.
